Running an open experiment: transparency and reproducibility in soil and ecosystem

### 2016-06-28 manuscript_openexperiment/Response_to_reviewers.Rmd

---
title: "Response_to_reviewers"
author: "Ben Bond-Lamberty"
date: "June 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We thank all referees for their thoughtful comments and careful assessment. Responses to their comments (italicized) are below, and significant changes in the manuscript text are marked in red.

@@

*As a journal editor, I have been pushing in every article to provide a clear data access statement and to publish data in open and public repositories. It's nice to have a manuscript to point to on good reason for doing so with a clear example.*

Thank you.

*I don't have any recommendations for major changes to this short piece. However, I think the caveats section (section 3) could benefit from more discussion on barriers to adopting this approach, not just from the incentive side, but from the skills needed, time invested, and challenges in using systems like github to run experiments. It would be helpful to discuss what was hard in getting the open experiment to work, what took the most time, what tools were the most difficult or least developed or accessible to scientists. If anything most of us are well convinced of the arguments the authors make on why to share data and code and incentives are showing up in multiple places. The challenge is finding the time, motivation, and technical skills to do so. So perhaps some discussion of how the authors dealt with those would help convince the otherwise skittish researcher that this isn’t as hard as it looks to do!*

We have taken much of the material in the former "Conclusions" section and split into a new section, "Barriers and caveats", which discusses some of these issues (skills, time, barriers for an 'average' scientist) in much greater depth.

### 2016-06-28 manuscript_openexperiment/Open_experiment.Rmd
Responses to Reviewer 2; significant new text in ms re barriers and caveats.

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "July 1, 2016"
output:
  html_document: default
  word_document: default
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

#### 3. Caveats and conclusions

By itself transparency does not guarantee repeatability and reproducibility {Easterbrook, 2014 #4137}, and it may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.

@@

We also recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}. That is, scientists do not receive 'proper' credit for datasets relative to publications, even when the former may be much more broadly valuable, in almost every area: hiring, promotion and tenure decisions, impact metrics, etc. The advent of 'data descriptor' articles, in journals such as *Scientific Data*, improves but does not solve this problem.

The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. First, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. Second, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data. Third, Git and GitHub are designed for working with code, not data. Although this situation continues to improve on a technical level[^9], it still requires a fair amount of coding expertise to set up and run an experimental repository such as the one detailed here. We applaud efforts to lower the technical and information hurdles to open science {Hampton, 2015 #4420}.

[^9]: https://github.com/blog/1885-better-word-highlighting-in-diffs

In summary, a variety of forces continue to push scientists towards more transparency in their methods, code, and data, with the goals of increasing reproducibility, enabling syntheses and meta-analyses, and improving trust in, and return from, publicly-funded science. The open experiment example we highlight here offers instrument-to-final product reproducibility and a very high level of transparency, although it is only one of a number of possible models {Hampton, 2015 #4420}. Elements of this case study (e.g., the use of issue-tracking software) might be usefully adopted in isolation, but we hope the entire experiment will be an example of individual scientists' decisions and practices {Wolkovich, 2012 #3303} having a larger impact. We encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.

>>>>

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "July 1, 2016"
output:
  word_document: default
  html_document: default
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

#### 3. Barriers and caveats

There were thus significant advantages to implementing and using an automated analytical pipeline in an open repository. It is worth considering in depth, however, the costs of such an approach, the investments required, and the barriers we encountered (or that many other scientists would).

* More upfront time is required in an open repository system, as the basic building blocks are assembled into an automated pipeline that runs from raw data to diagnostics to final products. While frustrating, this is generally a good thing however: automation makes later analytical stages much faster (see above), while openness provides a strong incentive to write clean, clear code from the beginning, for example, as opposed to deferring documentation.
* To our knowledge, there is no 'template' for an experimental repository such as the one used here. This results in unnecessary time re-inventing wheels, in particular with respect to repository design decisions.
* Effective data management takes some programming skills. We believe that basic programming, like basic statistics, should be a part of almost every scientist's training, and do not view this as a huge problem going forward. For scientists used to spreadsheet-oriented data processing, however, it requires new skills and practices.
* Version control (here, the Git software) is critical for tracking provenance, ensuring robustness, and efficiently sharing changes. It introduces significant complexity, however, particularly for any use beyond the most simple operations. We believe version control software remains too difficult for most scientists to use effectively[^9].
* Git and GitHub are fundamentally designed for working with code, not data or long documents. This situation continues to improve on a technical level[^10], but some operations (e.g., storing very large files; tracking small changes to columnar data; commenting on words or phrases) remain awkward or even impossible.

[^9]: https://xkcd.com/1597/
[^10]: E.g., https://github.com/blog/1885-better-word-highlighting-in-diffs

In summary, while tools and platforms such as Git, GitHub, RStudio, and RMarkdown have drastically improved the feasibility and accessibility of running an 'open experiment', they remain significant hurdles for many scientists. We applaud efforts to lower the technical and information hurdles to open science {Hampton, 2015 #4420}.

A number of caveats are also in order. The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. For example, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. In addition, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data.

More generally, by itself transparency does not guarantee repeatability and reproducibility {Easterbrook, 2014 #4137}, and it may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.

@@

Finally, we recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}. That is, scientists do not receive 'proper' credit for datasets relative to publications, even when the former may be much more broadly valuable, in almost every area: hiring, promotion and tenure decisions, impact metrics, etc. The advent of 'data descriptor' articles, in journals such as *Scientific Data*, improves but does not solve this problem.

#### 4. Conclusions

A variety of forces continue to push scientists towards more transparency in their methods, code, and data, with the goals of increasing reproducibility, enabling syntheses and meta-analyses, and improving trust in, and return from, publicly-funded science. The open experiment example we highlight here offers instrument-to-final product reproducibility and a very high level of transparency, although it is only one of a number of possible models {Hampton, 2015 #4420}. Elements of this case study (e.g., the use of issue-tracking or version control software) might be usefully adopted in isolation, but we hope the entire experiment will be an example of individual scientists' decisions and practices {Wolkovich, 2012 #3303} having a larger impact. We encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.
